---
title: "Collecting plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Collecting plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A script that makes one plot can keep it in a variable. A script that makes forty
needs somewhere to put them, and the answer is usually a list. A list holds the
plots but nothing about them: not the size each one should be saved at, not the
folder it belongs in, and not the better title somebody asked for afterwards.

A plots collection holds all of that next to the plot, and keeps it *next to*
the plot rather than in it. Nothing you change is written into the ggplot, so
every change can be undone, and the same plot can be changed again and again
without being rebuilt.

```{r setup}
library(ggview)
library(ggplot2)
library(tibble)
```

## Start a collection and add to it

`plots_init()` starts one and `plots_append()` adds to it. The name is the
plot's key, and slashes in it make folders later on.

```{r}
weight <- ggplot(mtcars, aes(wt, mpg, color = factor(cyl))) +
  geom_point(size = 2) +
  labs(title = "Heavier cars use more fuel", subtitle = "N = 32", color = "Cylinders")

cylinders <- ggplot(mtcars, aes(factor(cyl))) +
  geom_bar() +
  labs(title = "Most of these cars are eight-cylinder")

plots <- plots_init(name = "Engine notes", description = "Two plots from mtcars")
plots <- plots_append(plots, weight, name = "Engine/weight", type = "point")
plots <- plots_append(plots, cylinders, name = "Engine/cylinders", type = "bar")
```

In RStudio each of those calls also previews the plot at its true output size,
and so does every `plots_set()` below. That is the `show` argument, on whenever
the session is interactive; pass `show = FALSE` when adding or tuning several
plots at once.

```{r}
plots
```

In RStudio each name in that listing is a link: click one and the plot opens in
the viewer. A plot carrying a change is marked with a `*`. `as_tibble()` gives
the plain tibble underneath.

```{r}
as_tibble(plots)
```

## What a plot lets you change

`plots_params()` answers that. It lists each parameter, what kind of value it
takes, what the plot does without a change, and what it has been changed to.

```{r}
plots_params(plots, "Engine/weight")
```

`plots_set()` records a change. The stored plot is not touched — the change sits
beside it, and `plots_get()` applies it on the way out.

```{r, fig.width = 7, fig.height = 4}
plots_get(plots, "Engine/weight")
```

```{r, fig.width = 7, fig.height = 4}
plots <- plots_set(
  plots, "Engine/weight",
  title = "Weight explains most of the mileage gap",
  caption = "Source: mtcars",
  width = 8, height = 5
)

plots_get(plots, "Engine/weight")
```

```{r}
# The plot itself still carries the title it was built with.
plots$plot[[1]]$labels$title
```

Three things you can do to a parameter, and they are all different:

```{r}
# Leave it alone: not mentioning it changes nothing.
# Set it:
plots <- plots_set(plots, "Engine/weight", subtitle = "All 32 cars")
# Take it away: NA means the plot should not have one.
plots <- plots_set(plots, "Engine/weight", subtitle = NA)
```

`plots_reset()` forgets changes and puts the canvas back to the default for the
plot's type. What comes back is what the plot always was.

```{r}
plots <- plots_reset(plots, "Engine/weight", params = "subtitle")
plots_get(plots, "Engine/weight")$labels$subtitle
```

A change is checked before it is stored, so a typo fails where you made it
rather than when the plot is drawn:

```{r, error = TRUE}
plots_set(plots, "Engine/weight", color = "red")
```

## Going further than the default

`customizer_default()` offers the labels every plot has, and that is all it
claims to know. The plot above colors its points by cylinder count, and nothing
in that list lets anyone change those colors — the default cannot see them.

So say what this plot allows. A customizer is two functions. `apply` takes the
stored plot and the changed values and returns a new plot. `params` takes the
stored plot and says what may be changed, using the `param_*()` functions.
Keeping them together is what lets a program offer the right control for each
parameter without knowing anything about plots.

```{r}
recolor <- customizer(
  apply = function(plot, title = NULL, colors = NULL) {
    if (!is.null(title)) plot <- plot + labs(title = title)
    if (!is.null(colors)) {
      plot <- suppressMessages(plot + scale_color_manual(values = unlist(colors)))
    }
    plot
  },
  params = function(plot) {
    # Ask the plot what it colors by, rather than assuming. ggplot2 registers the
    # aesthetic as `colour` however you spell it in `aes()`.
    built <- suppressMessages(ggplot_build(plot))
    levels <- built$plot$scales$get_scales("colour")$range$range
    list(
      title  = param_text("Title", default = plot$labels$title),
      colors = param_mapping("Colors", keys = levels, to = "color")
    )
  }
)
```

Two things make this worth the trouble. The choices come from the plot's own
data, so a rerun with new categories offers the new ones. And `apply` may
rebuild a plot rather than add to it, so a customizer is not limited to what
`ggplot2` lets you add on top — filtering categories or reordering them is
possible here, and impossible with a layer.

Give the same plot the new customizer by adding it again under the same name. It
keeps its place and its id, and re-adding re-declares the plot, so it starts with
no changes recorded.

```{r}
plots <- plots_append(plots, weight, name = "Engine/weight", type = "point",
                      customizer = recolor)

plots_params(plots, "Engine/weight")
```

Four parameters where there were eight — this customizer offers a title and not
the rest of the labels — but one of them is a color per cylinder count, which the
default could never reach.

```{r, fig.width = 7, fig.height = 4}
plots <- plots_set(plots, "Engine/weight",
                   colors = c("4" = "#36c8ef", "6" = "#50bd90", "8" = "#de425b"))

plots_get(plots, "Engine/weight")
```


A customizer must survive being read back somewhere else, so it may not reach
for anything that only exists in the session that wrote it:

```{r, error = TRUE}
house_palette <- c("#36c8ef", "#50bd90")

customizer(
  apply = function(plot, a = NULL) plot + scale_fill_manual(values = house_palette),
  params = function(plot) list(a = param_text("A"))
)
```

Pass it in as a parameter instead, with the palette as the default.

## What a program reads

`plots_params()` returns plain data, one entry per parameter, which is what a
user interface is built from. `type` decides the control.

| `type` | What it takes | Control |
|---|---|---|
| `text` | one string | a box |
| `number` | one number, with optional `min`, `max`, `step` | a spinner |
| `flag` | `TRUE` or `FALSE` | a switch |
| `color` | one color | a picker |
| `choice` | one of `choices` | a list |
| `choices` | several of `choices` | a multiple list |
| `mapping` | a named value per key in `keys`, of the type it maps `to` | one control per key |

```{r}
str(plots_params(plots, "Engine/weight")$colors)
```

`default` is what the plot does on its own, `value` is what somebody changed it
to, and `value` is absent when nobody has. So the state to show is the value if
there is one, and the default if there is not — and a reset is dropping the
value.

Every plot also accepts `width` and `height`, whatever its customizer offers.
The canvas belongs to the collection rather than to any customizer, which is why
a customizer may not offer those two names.

## Sizing

Because the sizes are ordinary columns, re-sizing every plot of one type is one
line:

```{r}
library(dplyr)

plots <- plots |> mutate(height = if_else(type == "bar", 6, height))
plots
```

Each type starts from a default you can set for the whole collection.

```{r}
plots_dims(heatmap = c(18, 11))
```

## Save the collection

`save_plots()` writes one file per plot, each at its own size, and makes the
folders in the names.

```{r}
dir <- file.path(tempdir(), "vignette-plots")
files <- save_plots(plots, dir = dir)

list.files(dir, recursive = TRUE)
```

`as.list()` gives the same plots as a named list, ready for any other way of
writing them.

```{r}
names(as.list(plots))
```

## Store it and pick it up again

A collection is an ordinary R object, so it survives a save and a read, changes
and all. `plots_meta()` says what the collection is.

```{r}
file <- tempfile(fileext = ".rds")
saveRDS(plots, file)

back <- readRDS(file)
str(plots_meta(back))
plots_params(back, "Engine/weight")$colors$value
```

One thing is worth knowing before storing a large collection. A ggplot keeps the
environment it was built in, and saving the plot saves that environment too.
Plots built at the top level of a script cost little, because the global
environment is stored by reference: twenty plots over a 4,000 row data frame
come to under 3 MB. A plot built inside a function instead carries that
function's local objects, so a builder function that holds a large object in
scope makes a large file.

## Browse the whole collection

`as.gallery()` turns a collection into a gallery, with the slashes in the names
as its folders. Printing it opens a browsable page in the RStudio viewer.

```{r, eval = FALSE}
as.gallery(plots)
```
